---
title: "Rstats for Researchers"
author: "Tyson S. Barrett"
date: "`r Sys.Date()`"
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook: default
documentclass: book
link-citations: yes
---

```{r, echo=FALSE}
## Run but not shown
## Getting data ready for the examples
library(foreign)
library(dplyr)
dem_df <- read.xport("~/Box Sync/GitHub/blog_rstats/assets/Data/NHANES_demographics_11.xpt")
med_df <- read.xport("~/Box Sync/GitHub/blog_rstats/assets/Data/NHANES_MedHeath_11.xpt")
men_df <- read.xport("~/Box Sync/GitHub/blog_rstats/assets/Data/NHANES_MentHealth_11.xpt")
act_df <- read.xport("~/Box Sync/GitHub/blog_rstats/assets/Data/NHANES_PhysActivity_11.xpt")
names(dem_df) <- tolower(names(dem_df))
names(med_df) <- tolower(names(med_df))
names(men_df) <- tolower(names(men_df))
names(act_df) <- tolower(names(act_df))
df <- dem_df %>%
  full_join(med_df, by="seqn") %>%
  full_join(men_df, by="seqn") %>%
  full_join(act_df, by="seqn")
```



# Chapter 5: Generalized Linear Models {-}

Generalized Linear Models (GLM's) are extensions of linear regression to areas where assumptions of normality and homoskedasticity do not hold. There are several versions of GLM's, each for different types and distributions of outcomes. We are going to go through several of the most common.

## Logistic Regression {-}





## Poisson Regression {-}





## Beta Regression {-}






## Two-Part Models {-}






## Other Useful Models {-}




```{r}

```



